Waveform Inversion via Source Extension

### 2025-05-29 repro/papers/ssevelan/slides/slidesSEG19.tex
prep for JV talk

\AtBeginSection[]{
{\setbeamertemplate{logo}{}
 \begin{frame}<beamer>
    \frametitle{Agenda}
    \tableofcontents[currentsection]
  \end{frame}
}
}
% Images
% Macros
\newcommand{\peq}{\,+\hspace{-0.15cm}=}
\newcommand{\bm}{\bar{m}}
\newcommand{\dm}{\delta m}
\newcommand{\dom}{\delta \bar{m}}
\newcommand{\doma}{\delta \bar{m}_{\alpha}}
\newcommand{\om}{\bar{m}}
\newcommand{\bH}{{\bf H}}
\newcommand{\bR}{{\bf R}}
\newcommand{\bcF}{{\bar{\cal F}}}
\newcommand{\cF}{{\cal F}}
\newcommand{\oF}{\bar{F}}
\newcommand{\of}{\bar{f}}
\newcommand{\ou}{\bar{u}}
\newcommand{\odF}{\bar{F}^{\dagger}}
\newcommand{\oS}{\bar{S}}
\newcommand{\oM}{\bar{M}}
\newcommand{\oJa}{\bar{J}_{\alpha}}
\newcommand{\Na}{N_{\alpha}}
\newcommand{\dd}{\delta d}
\newcommand{\ddl}{\delta d_{\alpha}}
\title[]{Waveform Inversion via Source Extension}
\author[]{William W. Symes \inst{1}}
\institute[]{\inst{1} Rice University}
\date{Workshop in honor of Jean Virieux \\June 2025}

@@

\section{Inversion}

\begin{frame}%\frametitle{Conventional FWI}
FWI: adjust model parameters to make predicted data close to
observed data.

Predicted data = $Ru$, $u$ = modeled wavefield in space-time, $R$ =
sampling operator. Depends on $m$ = model parameters through wave equation
\[
L[m]u = s 
\]
$L[m]$ = wave operator, $s$ = energy source wavefield, Predicted data: $= F[m]s$, $F[m] = RL[m]^{-1}$ =
modeling operator

Least squares (``FWI''): given observed data $d$, source $s$, find $m$ to minimize
\[
J_{\rm FWI}[m] = \sum (F[m]s - d)^2
\]
$\sum$ = sum over all sources, receivers, times
\end{frame}

\begin{frame}
How to do it?

\begin{itemize}
\item Modeling (computing $u$ etc.): many methods - e.g. time-domain
  staggered grid finite differernce (\textcolor{blue}{Virieux}, 84)
\vspace{0,5cm}
\item Iterative descent update using gradient $\nabla J_{\rm FWI}$ - efficient
  computation via {\em adjoint state method} (Chavent-Lemmonier
  74,...), first multi-D demo by Gauthier-\textcolor{blue}{Virieux}-Tarantola 86
\end{itemize}

So what happens?
\end{frame}

\begin{frame}
A simple example: acoustic lens with isotropic point source @ $\bx_s$:
$f = w(t)\delta(\bx-\bx_s)$, $\bx_s$ = source position\\
%\[
%\frac{1}{\kappa}\frac{\partial p}{\partial t} = -\nabla \cdot {\bf v} + w(t)\delta(\bx-\bx_s), \,\rho \frac{\partial {\bf v}}{\partial t} = - \nabla p; \,p,{\bf v}=0, t \ll 0
%\]
%\vspace{-1.25in}
\vspace{-0.25in}
%\hspace{-0.5in}\includegraphics[height=4.0in]{Fig/bml0c.pdf}\\
%\includegraphics[height=2.0in}{Fig/bml0.pdf}\\
\begin{center}
\vspace{0.25in}
\includegraphics[height=1.25in]{Fig/bml0.png}\includegraphics[height=1.25in]{Fig/ptpwindl0.pdf}\\
\vspace{0.5cm} Left: target bulk modulus $\kappa$, source at
$x_s=3500$ m, $z_s=3000$ m, receivers at $x_r=2000-6000$ m, $z_r=1000$
m. Right: synthetic data, source pulse = [1.0, 2.0, 7.5, 12.5] Hz trapzoidal bandpass filter]
\end{center}
\end{frame}

%\begin{frame}
%Modeling operator $(S[\kappa]f)(t,x_r) = \phi(x_r) p(t,x_r,z_r)$ - linear in $f$, nonlinear in $\kappa$
%Model $m=[\kappa]$, wavefield $u=[p,\bv]$, $L[m]u=$ acoustic wave system, $R$ = sample $p$ at receivers, 
%\begin{center}
%\includegraphics[height=1.5in]{Fig/pulse00.pdf} \includegraphics[height=1.5in]{Fig/ptpwindl0.pdf}\\
%\vspace{0.25in}
%Left: [1.0, 2.0, 7.5, 12.5] Hz trapzoidal bandpass filter pulse $w$; Right: pressure traces $Ru$ at $z_r=1000 m$ for source at $x_s=3500 m, z_s=3000 m$. 
%\end{center}
%\end{frame}

\begin{frame}
\begin{center}
\includegraphics[height=1.25in]{Fig/covmestfwi2.png}\includegraphics[height=1.25in]{Fig/residcovmestfwi2wind.pdf}\\
\vspace{0.5cm}
Left: target model; 
Right: FWI estimate of $m$ = bulk modulus: initial model $\kappa_0
\equiv 4.0$ GPa, 450 steps steepest descent with smoothing
preconditioner. Right: data residual ($Ru - d$) - energy (norm)
$\approx 25$\% of data norm.
\end{center}

Failure mechanism: ``cycle skipping'', first documented in Gauthier-
\textcolor{blue}{Virieux}-Tarantola 86 - half-wavelength
criterion, (\textcolor{blue}{Virieux}-Operto 09)
\end{frame}

%\begin{frame}
%\begin{center}
%\includegraphics[height=1.2in]{Fig/ptpwindl0.pdf}\includegraphics[height=1.2in]{Fig/fwitrreslh0wind.pdf}\\
%\vspace{0.5cm}
%Left: target data; 
%Right: data residual after 20 KGN steps - final RMS residual reduction = 0.26
%\|d\|=9.72 \|S-d\|=2.51
%\end{center}
%\end{frame}

%\begin{frame}
%Big problem: "cycle skipping"
%\begin{itemize}
%\item $J[m]$ appears to have many (local) minima
%\item to avoid getting stuck at wrong estimates, initial $m$ (and $f$) must predict eve%nt times in $u$ ``correct within a wavelength''
%\end{itemize}

%OK with sufficiently low frequency data with high S/N, but not always available
%\end{frame}

\section{Extension}
\begin{frame}
  Many proposed cures for cycle skipping:
  \begin{itemize}
  \item acquire lower frequency data (Dellinger et al SEG 2016,...) 
  \item combine with traveltime tomography (Luo-Schuster 91,
    Prieux-\ldots-\textcolor{blue}{Virieux} 13)
  \item use a different objective, e.g. transport metric
    (Yang-Engquist 18, M\'{e}tivier-...-\textcolor{blue}{Virieux} 18)
  \item use a different model - {\em enlarge search space, easier data
    fit} (``model extension'')
  \end{itemize}

Many extension methods - Operto et al. 23, Farshad-Chauris 23, Huang
et al 19, S. 08. Common theme: maintain data fit while suppressing
additional degrees of freedom
\end{frame}

\begin{frame}
  \vspace{0.25in}
Example: Adaptive Waveform Inversion (AWI, Warner-Guasch 14): acoustic
model, {\em known} isotropic point source $s = w(t)\delta(\bx-\bx_s)$,

Idea: if events in predicted data $F[m]s$ are in wrong place, use a 
filter $f$ to put them in the right place: extended modeling op $\bar{F}[m,f] = f*F[m]$.
\vspace{-0.25in}
\begin{center}
  \includegraphics[height=1in]{Fig/ptph0half.pdf}\includegraphics[height=1in]{Fig/uest0half.pdf}\includegraphics[height=1in]{Fig/ptpl0half.pdf}\\
\vspace{0.5cm}
Left: Predicted data $F[m_0]s$, $m_0=$ homogenous model; Center: adaptive filter $f$;
Right: adapted data $f*F[m_0]s \approx d =$ lens data
\end{center}

\end{frame}
\begin{frame}
$f$ = $d$ deconvolved (trace-by-trace) by $F[m]s$ (``easy''), so can
make $\bar{F}[m,f]s$ fit data {\em for any m} (common feature of extension methods)

but $\bar{F}[m,f] \ne F[m]$ unless $f(\bx_s,\bx_r,t)=\delta(t)$ - to
test, find an operator vanishing on $\delta(t)$ (``annihilator'' -
another common feature), update $m$ to reduce its output

$t\delta(t)=0$, so minimize
\[
J_{\rm AWI}[m] = \sum_{\bx_s,\bx_r}\frac{{\sum_t|\color{blue}t}
  f[m](\bx_s,\bx_r,t)|^2}{\sum_{t}| f[m](\bx_s,\bx_r,t)|^2}
\]
\end{frame}

%\begin{frame}
%\vspace{-1in}
%\begin{center}
%\hspace{-0.5in}\includegraphics[height=2.5in]{Fig/uest0gz1800wind.pdf}\hspace{-0.5in}\includegraphics[height=2.5in]{Fig/uestmswigz1800wind.pdf}\\
%\vspace{-0.75in}
%Left: $f[m]$, $m$=initial model. Right: $f[m]$, $m$ updated to reduce
%$J_{\rm AWI}$
%\end{center}
%\end{frame}

\begin{frame}
  Good News: suppose
  \begin{itemize}
  \item transmission data dominated by single arrivals - unique
    travel time $\tau$ between source and receiver,
  \item $d \approx F[m_*]s$
  \item RMS wavelength of $w \rightarrow 0$.
  \end{itemize}
  Then
  \[
    J_{\rm AWI}[m] \rightarrow \sum_{\bx_s,\bx_r} |\tau[m]-\tau[m_*]|^2
  \]

  \textcolor{blue}{For single-arrival data, AWI is asymptotic to
    travel time tomography} $\Rightarrow$ no cycle-skipping

@@

  \section{Triplication}

\begin{frame}
  Conclusion does not apply to acoustic lens example - multiple
  arrivals.

  Asymptotic analysis, numerical examples $\Rightarrow$

  Significant energy in $d$ at multiple arrival times $\Rightarrow$
  approach based on source-receiver extension (e. g. AWI) behaves like
  FWI (S. 94, Plessix et al. 00, Huang et al. 17, Yong et al. 23.)

\begin{center}
\hspace{-0.5in}\includegraphics[height=0.8in]{Fig/uest0wind.pdf}\includegraphics[height=0.8in]{Fig/uestmswiwind.pdf}\\

Adaptive filters for lens data. Left: $f[m]$, $m$=initial model. Right: $f[m]$, $m$ updated to reduce
$J_{\rm MSWI}$ (similar to $J_{\rm AWI}$).
\end{center}
\end{frame}

\begin{frame}
  Remedy: localize inversion around individual events

  \begin{itemize}
  \item localize in time - Yong-Brossier-Métivier-\textcolor{blue}{Virieux} 23: Local
    Adaptive Waveform Inversion (LAWI) via Gabor filter
  \item localize in phase space - surface source extension (SSE) -
    Huang et al. 19
  \end{itemize}

  SSE extended source: $\bar{s} = \bar{w}(\bx,t;\bx_s)\delta_{{\cal
      S}}(\bx)$, ${\cal S} $= surface containing source locations -
  {\em spread source energy over a surface}, acts as antennae,
  controls full 2D phase spectrum

For any $m$, determine $\bar{w}=\bar{w}[m]$ so that $F\bar{s} \approx d$. Update $m$ to drive
$\bar{w}(\bx,t;\bx_s) \rightarrow w(t)\delta(\bx-\bx_s)$ by minimizing
\[
  J_{\rm SSE}[m] = \sum_{\bx_s}\sum_{\bx,t} ||(\bx-\bx_s)|
  \bar{w}[m](\bx,t;\bx_s)|^2
 \]
\end{frame}

%\begin{frame}
%Detailed Formulation:

%At solution, must recover known constraints on model (``get closer to Vancouver's latitude as you get close to the sea''):

%$\Rightarrow$ introduce {\color{blue} annihilator} $A$, vanishing on models obeying these constraints - minimize output jointly with data residual

@@

%Many choices. For SSE: scale by offset 
%\[
%A\of(\bx,t;x_s,y_s) = \sqrt{(x-x_s)^2+(y-y_s)^2}w(x,y,t;x_s,y_s)
%\]

%{\color {blue} Extension Principle:} $m,\of$ so  that $R\bar{u} \approx d$, $A\of \approx 0$ $\Rightarrow$ approx FWI solution
%\end{frame}

@@

\begin{frame}
Penalty method: minimize
\[
J_{\alpha}[m,\bar{w}] = \frac{1}{2} \sum |F[m]\bar{w} - d|^2 + \alpha^2||\bx-\bx_s|\bar{w}|^2
\]
over $m$

\begin{itemize}
\item time-domain FD, gradient via adjoint state
\item Variable Projection:
  \begin{itemize}
  \item update $\bar{w}$ using Conjugate Gradient Iteration (inner loop)
\item update $m$ using steepest descent, line search (outer loop) plus
  smoothing regularization
  \end{itemize}
\item update $\alpha$ using Discrepancy Principle (keep data residual $\in (e_-, e_+)$) - {\color{blue} can always choose $\bar{w}$ to fit data} $\Rightarrow$ can start with $\alpha=0$ (Fu \& S. 17)
\end{itemize}
\end{frame}

\begin{frame}
\begin{center}
SSE for 1 shot lens data, source surface ${\cal S} = \{z = 3000 m\}$
\end{center}
\begin{center}
\includegraphics[height=1.25in]{Fig/cgw1_bulkupdlh0.png}\includegraphics[height=1.25in]{Fig/cgw41_bulkupdlh0.png}\\
\vspace{0.25in}
Start at $\kappa_0=4.0$ GPa, use $e_-=0.05\|d\|, e_+=0.1\|d\|$. \\
Left: Iteration 1 ($\alpha=2.9 e-6$)\\
Right: Iteration 41 ($\alpha = 1.99 e-5$)
\end{center}
\end{frame}

\begin{frame}
$m=[\kappa]$ converges $\leftrightarrow$ $\alpha$ increases, extended source $\of$ (initially spread over $z=z_s=3000$ m) {\color{blue} focuses} at $x=x_s=3500$ m.
\begin{center}
\hspace{-1cm}\includegraphics[height=1.25in]{Fig/cgw0_est_source_plh0.pdf}\includegraphics[height=1.25in]{Fig/cgw8_est_source_plh0.pdf}\includegraphics[height=1.25in]{Fig/cgw18_est_source_plh0.pdf}\includegraphics[height=1.25in]{Fig/cgw40_est_source_plh0.pdf}\\
\vspace{0.25in}
Left to right: $\of$ at Iterations 1 ($\alpha=2.9 e-6$), 9
($\alpha= 5.2 e-6$ m) 19 ($\alpha=1.32 e-5$) and 41 ($\alpha=1.99e-5$).
\end{center}
\end{frame}

\begin{frame}
\begin{center}
\includegraphics[height=1.25in]{Fig/covmestfwicgw41.png}\includegraphics[height=1.25in]{Fig/residcovmestfwicgw41wind.pdf}\\
\vspace{0.25in}
Left: ``Jump to $\alpha=\infty$": bulk modulus estimated by FWI, 200
steepest descent steps starting with iteration 41 of SSE., final RMS residual reduction = 0.05;
Right: FWI data residual 
\end{center}
\end{frame}

\begin{frame}
  Why it works - a hint:

provided that
\begin{itemize}
\item $d = F[m_*]s$,
\item RMS wavelength of $w$ (hence $d$) $\rightarrow 0$,
\end{itemize}
\[
  \nabla \nabla  J_{\rm SSE}[m_*] \rightarrow \sum_{\bx,\bp}
\]
\end{frame}

>>>>

\AtBeginSection[]{
{\setbeamertemplate{logo}{}
 \begin{frame}<beamer>
    \frametitle{Agenda}
    \tableofcontents[currentsection]
  \end{frame}
}
}
% Images
% Macros
\newcommand{\peq}{\,+\hspace{-0.15cm}=}
\newcommand{\bm}{\bar{m}}
\newcommand{\dm}{\delta m}
\newcommand{\dom}{\delta \bar{m}}
\newcommand{\doma}{\delta \bar{m}_{\alpha}}
\newcommand{\om}{\bar{m}}
\newcommand{\bH}{{\bf H}}
\newcommand{\bR}{{\bf R}}
\newcommand{\bcF}{{\bar{\cal F}}}
\newcommand{\cF}{{\cal F}}
\newcommand{\oF}{\bar{F}}
\newcommand{\of}{\bar{f}}
\newcommand{\ou}{\bar{u}}
\newcommand{\odF}{\bar{F}^{\dagger}}
\newcommand{\oS}{\bar{S}}
\newcommand{\oM}{\bar{M}}
\newcommand{\oJa}{\bar{J}_{\alpha}}
\newcommand{\Na}{N_{\alpha}}
\newcommand{\dd}{\delta d}
\newcommand{\ddl}{\delta d_{\alpha}}
\title[]{Waveform Inversion via Source Extension}
\author[]{Guanghui Huang \inst{1} \and Rami Nammour \inst{2} \and \\William W. Symes \inst{3} \and Mohamed Dolliazal \inst{2}}
\institute[]{\inst{1} Michigan State University \and \inst{2} Total $E\&P$ $R\&T$ USA \and \inst{3} Rice University}
\date{September 2019}

@@

\section{What is ``extended modeling'', and why bother}

\begin{frame}%\frametitle{Conventional FWI}
FWI: adjust model to minimize data misfit (observed - modeled)

Wave equation:
\[
L[m]u = f 
\]
$m$ = model parameters, $L$ = wave operator, $u$ = modeled wavefield(s), $f$ = energy source model

FWI objective: 
\[
J_{\rm FWI}[m] = \sum (R u - d)^2
\]
$R$ = sampling operator, $Ru$ = modeled data, $d$ = observed data, $\sum$ = sum over all components

Note: Wave equation $\Rightarrow$ $u = u[m,f]$
\end{frame}

\begin{frame}
A simple example: acoustic transmission inversion w/ isotropic point source @ $\bx_s$: $f = w(t)\delta(\bx-\bx_s)$
%\[
%\frac{1}{\kappa}\frac{\partial p}{\partial t} = -\nabla \cdot {\bf v} + w(t)\delta(\bx-\bx_s), \,\rho \frac{\partial {\bf v}}{\partial t} = - \nabla p; \,p,{\bf v}=0, t \ll 0
%\]
\vspace{-1.25in}
%\vspace{-0.5in}
\begin{center}
\hspace{-0.5in}\includegraphics[height=4.0in]{Fig/bml0c.pdf}\\
%\includegraphics[height=2.0in}{Fig/bml0.pdf}\\
\vspace{-1in}
target bulk modulus $\kappa$, source at $x_s=3500$ m, $z_s=3000$ m, receivers at $x_r=2000-6000$ m, $z_r=1000$ m
\end{center}
\end{frame}

\begin{frame}
%Modeling operator $(S[\kappa]f)(t,x_r) = \phi(x_r) p(t,x_r,z_r)$ - linear in $f$, nonlinear in $\kappa$
Model $m=[\kappa]$, wavefield $u=[p,\bv]$, $L[m]u=$ acoustic wave system, $R$ = sample $p$ at receivers, 
\begin{center}
\includegraphics[height=1.5in]{Fig/pulse00.pdf} \includegraphics[height=1.5in]{Fig/ptpwindl0.pdf}\\
\vspace{0.25in}
Left: [1.0, 2.0, 7.5, 12.5] Hz trapzoidal bandpass filter pulse $w$; Right: pressure traces $Ru$ at $z_r=1000 m$ for source at $x_s=3500 m, z_s=3000 m$. 
\end{center}
\end{frame}

\begin{frame}
Minimizing $J_{\rm FWI}[m] = \sum (R u - d)^2$:

\begin{itemize}
\item Modeling (computing $u$ etc.): time-domain Finite Difference
\vspace{0,5cm}
\item Efficient comp'n of $\nabla J_{\rm FWI}$ = ``adjoint state method'' (Chavent-Lemmonier 74,...) - $\sim$ RTM
\vspace{0.5cm}
\item Iteratively update $m$ to reduce $J_{\rm FWI}$ - Newton's or related gradient/descent method, plus smoothing regularization
\end{itemize}
\end{frame}

\begin{frame}
\begin{center}
\includegraphics[height=1.25in]{Fig/bml0.pdf}\includegraphics[height=1.25in]{Fig/fwitrbulkmodinvlh0.pdf}\\
\vspace{0.5cm}
Left: target model; 
Right: FWI estimate of $m$ = bulk modulus: initial model $\kappa_0 \equiv 4.0$ GPa, 20 iterations trust region Krylov-Gauss-Newton with smoothing preconditioner. Each KGN step =  up to ``inner'' 20 CG steps $\approx$ 400 fwd models
\end{center}
\end{frame}

\begin{frame}
\begin{center}
\includegraphics[height=1.2in]{Fig/ptpwindl0.pdf}\includegraphics[height=1.2in]{Fig/fwitrreslh0wind.pdf}\\
\vspace{0.5cm}
Left: target data; 
Right: data residual after 20 KGN steps - final RMS residual reduction = 0.26
%\|d\|=9.72 \|S-d\|=2.51
\end{center}
\end{frame}

\begin{frame}
Big problem: "cycle skipping"
\begin{itemize}
\item $J[m]$ appears to have many (local) minima
\item to avoid getting stuck at wrong estimates, initial $m$ (and $f$) must predict event times in $u$ ``correct within a wavelength''
\end{itemize}

OK with sufficiently low frequency data with high S/N, but not always available
\end{frame}

\begin{frame}\frametitle{An Analogy}
%\vspace{0.5cm}
Task: Given latitude of Vancouver ($49^{\circ} 15' N$), find its longitude by traveling there from somewhere in Canadian Rockies ($\approx 121^{\circ} W$)

Algorithm: start at correct latitude (eg. Sunshine Valley, $49^{\circ} 16' N$, $121^{\circ} 11' W$) and go {\color{blue} downhill and due west} until altitude minimized 
\vspace{-3cm}
\begin{center}
\includegraphics[width=0.6\textwidth,angle=270]{Fig/bc.pdf}
\end{center}
\vspace{-3cm}
\footnotetext{\tiny{https://www.google.com/maps/@49.2332511,-122.8181263,9z, 2019.08.18}}
\end{frame}

\begin{frame}
You, looking west. Now, {\color{blue} go downhill and due west}... oops.
\vspace{-1.5cm}
\begin{center}
\includegraphics[width=0.6\textwidth,angle=270]{Fig/mtns.pdf}
\end{center}
\vspace{-1cm}
%\tiny{{https://www.dreamingandwandering.com/2016/01/10/25-photos-of-the-canadian-rockies-that-will-make-you-pack-your-bags-and-go/}, 2019.08.21}
\footnotetext{\tiny{\noindent https://www.dreamingandwandering.com/2016/01/10/25-photos-of-the-canadian-rockies-that-will-make-you-pack-your-bags-and-go/}, 2019.08.21}
\end{frame}

\begin{frame}
Analogy w/ FWI: altitude $\sim$ $J_{\rm FWI}[m] =$ mean square error

Can't reduce to desired min = 0 while going downhill ($\sim$ using descent method) {\color{blue} and} maintaining latitude of Vancouver ($\sim$ known physical properties of solution)

Descent methods - mandatory because of computational size, can't drop

Alternative: drop ``physical constraint'' (latitude), search in larger ``unphysical'' space 
\end{frame}

\begin{frame}
With full 2D search available, can go to local creek, then float down to local river, join Fraser River near Hope ($49^{\circ} 23' N$) ... float your boat to Vancouver, downhill all the way!
\vspace{-2.8cm}
\begin{center}
\includegraphics[width=0.5\textwidth]{Fig/fraser.pdf}
\end{center}
\vspace{-3cm}
Moral: dropping some known properties (physical constraints) of solution {\color{blue} during search} may make descent search feasible

\footnotetext{\tiny{https://images.search.yahoo.com/yhs/search?p=fraser+river}, 2019.08.21}

@@

\section{FWI via source extension}
\begin{frame}

{\color{blue} Extended modeling:} allow interim models to violate known ``physical'' constraints on true solution

Active source data (usually) involves {\em localized} source model with known location, eg. ($\bx=(x,y,z)$)
\[
f(\bx,t;x_s,y_s) = w(t) \delta(\bx-\bx_s) = \mbox{ isotropic point source at }\bx=\bx_s
\]
{\color{blue} Source Extension}: replace $f$ with {\em non-localized} source model $\of$

Extended wavefield $\ou$  solves $L[m]\ou = \of$, data prediction = $R\ou$ - computational complexity similar to FWI

\begin{frame}
Source extension methods:
\begin{itemize}
\item
Wavefield Reconstruction Inversion (WRI):
van Leeuwen and Herrmann GJI 2013, Inv. Probs. 2016, Wang \& Yingst SEG 2016, Aghamiry et al GEO 2019 
\item Adaptive Waveform Inversion (AWI): Warner \& Guatsch, GEO 2014, 2016
\item ... (see abstract)
\item {\color{blue} surface source extension (SSE):} $\of(\bx,t;x_s,y_s) = w(x,y,t; x_s,y_s)\delta(z-z_s)$
\end{itemize}

Common feature: additional source degrees of freedom permit good fit to data {\em for large set of $m$}
\end{frame}

\begin{frame}
Detailed Formulation:

At solution, must recover known constraints on model (``get closer to Vancouver's latitude as you get close to the sea''):

$\Rightarrow$ introduce {\color{blue} annihilator} $A$, vanishing on models obeying these constraints - minimize output jointly with data residual

@@

Many choices. For SSE: scale by offset 
\[
A\of(\bx,t;x_s,y_s) = \sqrt{(x-x_s)^2+(y-y_s)^2}w(x,y,t;x_s,y_s)
\]

{\color {blue} Extension Principle:} $m,\of$ so  that $R\bar{u} \approx d$, $A\of \approx 0$ $\Rightarrow$ approx FWI solution
\end{frame}

@@

\begin{frame}
FWI via SSE: given $d$ and initial $m_0$, use numerical optimization method to find minimizer $m$ of
\[
J_{\rm SSE}[m] = \mbox{ min}_{\of}\frac{1}{2} \sum (R\ou - d)^2 + \alpha^2 \sum (A\of)^2
\]
over $m$

\begin{itemize}
\item time-domain FD, gradient via adjoint state
\item update $\of$ using Conjugate Gradient Iteration (inner loop)
\item update $m$ using steepest descent, line search (outer loop) plus smoothing regularization
\item update $\alpha$ using Discrepancy Principle (keep data residual $\in (e_-, e_+)$) - {\color{blue} can always choose $\of$ to fit data} $\Rightarrow$ can start with $\alpha=0$ (Fu \& S. GEO 2017)
\end{itemize}
\end{frame}

\begin{frame}
\begin{center}
Example 1: Lens model
\end{center}
\begin{center}
\includegraphics[height=1.25in]{Fig/cgw1bulkupdlh0.pdf}\includegraphics[height=1.25in]{Fig/cgw30bulkupdlh0.pdf}\\
\vspace{0.25in}
Start at $\kappa_0=4.0$ GPa, use $e_-=0.1\|d\|, e_+=0.2\|d\|$. \\
Left: Iteration 1 ($\alpha=4.83 \times 10^{-6}$)\\
Right: Iteration 30 ($\alpha = 9.1\times 10^{-5}$)
\end{center}
\end{frame}

\begin{frame}
$m=[\kappa]$ converges $\leftrightarrow$ $\alpha$ increases, extended source $\of$ (initially spread over $z=z_s=1000$ m) {\color{blue} focuses} at $x=x_s=3500$ m.
\begin{center}
\hspace{-1cm}\includegraphics[height=1.25in]{Fig/cgw1estsourceplh0.pdf}\includegraphics[height=1.25in]{Fig/cgw7estsourceplh0.pdf}\includegraphics[height=1.25in]{Fig/cgw29estsourceplh0.pdf}\\
\vspace{0.25in}
Left to right: $\of$ at Iterations 1 ($\alpha=4.83 \times 10^{-6}, r=2000$ m), 8 ($\alpha= 1.32 \times 10^{-5}, r=1000$ m) and 30 ($\alpha=9.1 \times 10^{-5}, r=250$ m).
\end{center}
\end{frame}

\begin{frame}
\begin{center}
\includegraphics[height=1.25in]{Fig/fwitrsse1bulkmodinvlh0.pdf}\includegraphics[height=1.25in]{Fig/bml0.pdf}\\
\vspace{0.25in}
Left: ``Jump to $\alpha=\infty$": bulk modulus estimated by FWI, starting with iteration 30 of SSE. 18 steps KGN, each $\le$ inner 20 CG steps, final RMS residual reduction = 0.012;
Right: Target bulk modulus field
\end{center}
\end{frame}
